\textls[-0]{\thetitle}

### 2023-03-22 canonicitymodel.tex
fixes

Now we prove canonicity (and consistency) for \TT using a logical-relations model.
We follow the reducibility argument of
\citet{kaposi2019gluing}, \citet{coquand2018canonicity}, and \citet{sterling2019algebraic}
to construct our model.
The meta\-languages of these prior models are based on QIITs, categories with
families \cite{dybjer1995internal}, and the generalized algebraic theory
\cite{cartmell1986generalised}, respectively.
Without exposing the reader to too many technical details, our meta\-language
should be understood an instance of any of the above logical frameworks---the difference is our formulation did quotient manually but if we use their framework, quotient is done automatically.
%Our model works directly on the earlier introduced syntax for readability\footnote{We still
%take quotient, so ultimately our syntax should be considered as an instance of
%any of the above logical framework, without exposing the reader too many
%technical details}.

@@

\noindent
Canonicity is a basic criterion that should be met by a dependent type theory serving as
a programming language or as a computational foundation for mathematics.
%\footnote{See nlab explanation: \href{https://ncatlab.org/nlab/show/canonical+form}{Canonical Form}}
% We can even argue that, if this theorem is proven in a computable meta\-logic, 
% then by the Curry–Howard correspondence, the canonicity theorem provides a big-step
% interpreter for closed terms of the boolean type.
%\YZ{Can we say something about
%strong/weak normalization?}\EDJreply{
%  No. \\
%  1. In the declarative system, we cannot talk about strong/weak normalization (because we don't have operational semantic, if you recall the definition of strong/weak normalization, it based on reduction/path of steps/op semantic). So any review says strong/weak normalization to you, please stop them. \\ 
%  2. In declarative system, we only say ``normalization'' directly, which means, we construct a function that \\
%      for arbitrary class of (judgemental-equivalent) terms, it will return one syntax representation (called normal-form) for each class \\
%    (apparently most people will make the normal form  without beta redex, however, interestingly, they may do eta expansion instead of eta reduction because theoretically easier) \\ 
%  3. So translation won't work because it compiles away the normal form of linkage (the $\mu^+$). This part requires a direct proof which construct normal form and neutral form for our calculus, and a normalization model from which we extract a runnable interpreter for open terms \\
%  4. Of course we can conjecture we have that normalization function \\
%  5. Also we do have close term normalization, which is canonicity. Or more operationally speaking, we have termination (a runnable interpreter) for close term (since QIIT is runnable, the canoncity model itself is the runnable interpreter), \\
%  but we don't have termination (a runnable interpreter) for open terms.  \\ 
%  So earlier you say ``normalize'' a recursion on an inductive type, that is no doable -- if the programmer wants to program on meta-theory, they will have to manually write out the a chain of judgemental equality (with a bunch of beta rules inside) by themselves. 
%}

We first need the mathematical setup to interpret universe levels, following \citet{sterling2019algebraic}:

@@

We can roughly consider each Grothendieck universe $\Set{i}$ as the $\Set{i}$ in Agda: 
\begin{itemize}
  \item Each $\Set{i}$ is closed under dependent function types and dependent
  pair types. For example, later, for our interpretation of dependent function
  types, when we have $\denotesC{A}, \denotesC{B} \in \Set{i}$, we will have
  $\denotesC{\TyPi{A}{B}} \in \Set{i}$.
  \item The universe hierarchy is cumulative, as $\Set{i} \in \Set{i+1}$ and $\Set{i} \subseteq \Set{i+1}$.
  \item Thus if $\denotesC{A} \in \Set{i}, \denotesC{B} \in \Set{j}$, we will
  have $\denotesC{\TyPi{A}{B}} \in \Set{i} \cup \Set{j} = \Set{i \lcup j}$.
\end{itemize}

@@

  \denotesC{\goodCtx{\Gm}{k}} \text{ is a function }&: \{\gm : \goodSub{\cdot}{\gm}{\Gm} \} \to \Set{k} \quad \text{ (also sets indexed by closed substitution)}\\

@@

  \denotesC{\goodType{\Gm}{T}{j}} \text{ is a dependent function }&: \prod_{\goodSub{\cdot}{\gm}{\Gm}}\prod_{\gm' \in \denotesC{\goodCtx{\Gm}{}}(\gm)}\{t : \goodTerm{\cdot}{t}{\sub{T}{\gm}}\}\to\Set{j} \\

@@

  \denotesC{\goodTerm{\Gm}{t}{T}} \text{ is a dependent function}&: \prod_{\goodSub{\cdot}{\gm}{\Gm}} \prod_{\gm' \in \denotesC{\goodCtx{\Gm}{}}}\denotesC{\goodType{\Gm}{T}{}}(\gm)(\gm')(\sub{t}{\gm})\\

@@

  \denotesC{\goodType{\Gm}{\TyPi{A}{B}}{}}(\gm)(\gm')(t) &= \prod_{\goodTerm{\cdot}{u}{\sub{A}{\gm}}} \ \ \prod_{u' \in \denotesC{A}(\gm)(\gm')(u)} \denotesC{B}(\SubstExt{\gm}{u})((\gm', u'))(\sub{\app{t}}{\SubstExt{"id"}{u}})\\

@@

  \denotesC{\goodTerm{\Gm}{\lam{t}}{\TyPi{A}{B}}}(\gm)(\gm') &= \lambda u \lambda u' . \denotesC{t}(\SubstExt{\gm}{u}) ((\gm' , u')) \\

>>>>

Now we prove canonicity (and consistency) for \TT using a logical-relations model.
We follow the reducibility argument of
\citet{kaposi2019gluing}, \citet{coquand2018canonicity}, and \citet{sterling2019algebraic}
to construct our model.
The meta\-languages of these prior models are based on QIITs, categories with
families \cite{dybjer1995internal}, and the generalized algebraic theory
\cite{cartmell1986generalised}, respectively.
Without exposing the reader to too many technical details, our meta\-language
should be understood as an instance of any of the above logical frameworks---the
difference is that quotienting is manual in our formulation, whereas it is
automatic with the logical frameworks.
%Our model works directly on the earlier introduced syntax for readability\footnote{We still
%take quotient, so ultimately our syntax should be considered as an instance of
%any of the above logical framework, without exposing the reader too many
%technical details}.

@@

\noindent
Canonicity is a fundamental property that should be satisfied by a dependent type theory serving as
a programming language or as a computational foundation for mathematics.
%\footnote{See nlab explanation: \href{https://ncatlab.org/nlab/show/canonical+form}{Canonical Form}}
% We can even argue that, if this theorem is proven in a computable meta\-logic, 
% then by the Curry–Howard correspondence, the canonicity theorem provides a big-step
% interpreter for closed terms of the boolean type.
%\YZ{Can we say something about
%strong/weak normalization?}\EDJreply{
%  No. \\
%  1. In the declarative system, we cannot talk about strong/weak normalization (because we don't have operational semantic, if you recall the definition of strong/weak normalization, it based on reduction/path of steps/op semantic). So any review says strong/weak normalization to you, please stop them. \\ 
%  2. In declarative system, we only say ``normalization'' directly, which means, we construct a function that \\
%      for arbitrary class of (judgemental-equivalent) terms, it will return one syntax representation (called normal-form) for each class \\
%    (apparently most people will make the normal form  without beta redex, however, interestingly, they may do eta expansion instead of eta reduction because theoretically easier) \\ 
%  3. So translation won't work because it compiles away the normal form of linkage (the $\mu^+$). This part requires a direct proof which construct normal form and neutral form for our calculus, and a normalization model from which we extract a runnable interpreter for open terms \\
%  4. Of course we can conjecture we have that normalization function \\
%  5. Also we do have close term normalization, which is canonicity. Or more operationally speaking, we have termination (a runnable interpreter) for close term (since QIIT is runnable, the canoncity model itself is the runnable interpreter), \\
%  but we don't have termination (a runnable interpreter) for open terms.  \\ 
%  So earlier you say ``normalize'' a recursion on an inductive type, that is no doable -- if the programmer wants to program on meta-theory, they will have to manually write out the a chain of judgemental equality (with a bunch of beta rules inside) by themselves. 
%}

First, we need the mathematical setup to interpret universe levels, following \citet{sterling2019algebraic}:

@@

We can roughly consider each Grothendieck universe $\Set{i}$ as the $\Set{i}$ in Agda: 
\begin{itemize}
  \item Each $\Set{i}$ is closed under dependent function types and dependent
  pair types. For example, later, for our interpretation of dependent function
  types, when we have $\denotesC{A}, \denotesC{B} \in \Set{i}$, we will have
  $\denotesC{\TyPi{A}{B}} \in \Set{i}$.
  \item The universe hierarchy is cumulative, as $\Set{i} \in \Set{i+1}$ and $\Set{i} \subseteq \Set{i+1}$.
  \item Thus, if $\denotesC{A} \in \Set{i}, \denotesC{B} \in \Set{j}$, we will
  have $\denotesC{\TyPi{A}{B}} \in \Set{i} \cup \Set{j} = \Set{i \lcup j}$.
\end{itemize}

@@

  \denotesC{\goodCtx{\Gm}{k}} \text{ is a function }&: \setc{\gm}{\goodSub{\cdot}{\gm}{\Gm}} \to \Set{k} \quad \text{ (i.e., sets indexed by closed substitution)}\\

@@

  \denotesC{\goodType{\Gm}{T}{j}} \text{ is a dependent function }&: \prod_{\goodSub{\cdot}{\gm}{\Gm}}\prod_{\gm' \in \denotesC{\goodCtx{\Gm}{}}(\gm)}\setc{t}{\goodTerm{\cdot}{t}{\sub{T}{\gm}}}\to\Set{j} \\

@@

  \denotesC{\goodTerm{\Gm}{t}{T}} \text{ is a dependent function}&: \prod_{\goodSub{\cdot}{\gm}{\Gm}} \prod_{\gm' \in \denotesC{\goodCtx{\Gm}{}}(\gm)}\denotesC{\goodType{\Gm}{T}{}}(\gm)(\gm')(\sub{t}{\gm})\\

@@

  \denotesC{\goodType{\Gm}{\TyPi{A}{B}}{}}(\gm)(\gm')(t) &= \prod_{\goodTerm{\cdot}{u}{\sub{A}{\gm}}} \ \ \prod_{u' \in \denotesC{A}(\gm)(\gm')(u)} \denotesC{B}(\SubstExt{\gm}{u})(\gm', u')(\sub{\app{t}}{\SubstExt{"id"}{u}})\\

@@

  \denotesC{\goodTerm{\Gm}{\lam{t}}{\TyPi{A}{B}}}(\gm)(\gm') &= \lambda u \lambda u' . \denotesC{t}(\SubstExt{\gm}{u}) (\gm' , u') \\

### 2023-03-22 syn-translate.tex
fixes

We will first define two helpers
\begin{itemize}
  \item a new type $\goodType{\Gamma}{\Sigr{\Gamma}{n}{j}}{}$  when
  $\goodCtx{\Gamma}{}$  is well-formed and $n \in \mathbb{N}$
  such that $\goodTerm{\Gm}{\denotesT{\goodSigU{j}{\Gamma}{\lsig}{n}}}{\Sigr{\Gm}{n}{j}}$
  \item and a function $\mathscr{P} : \setc{\lsig'}{ \goodTerm{\Gm}{\lsig'}{\Sigr{\Gm}{n}{j}} } \to \setc{T}{ \goodType{\Gm}{T}{} }$ mapping a term of type $\Sigr{\Gm}{n}{j}$ into a type.
\end{itemize}
They are defined mutually inductively on $n$, the signature length.

\begin{align*}
  {\Sigr{\Gamma}{(n+1)}{j}} &= 
    \Sigma~(\Sigr{\Gamma}{n}{j})
          ~(\Sigma~\cU_j \\
          & \quad \quad ~(\Sigma~(\TyPi{{\mathscr{P}({\sub{\var{0}}{\SubstWeak{1}}})}}{\El{\sub{\var{0}}{\SubstWeak{1}}}}) \\
          & \quad \quad \quad \TyPi{\El{\sub{\var{0}}{\SubstWeak{1}}}}{\cU_j})) \\
  \Sigr{\Gamma}{0}{j} &= \top \\
  \mathscr{P}(\lsig') &=  
    \Sigma~(\fst{\lsig'})~(\El{\app{"snd"^3~\lsig'}[(\SubstWeak{1}, \app{\fst{("snd"^2~{\lsig'})}})]}) \\ & \text{ when } \goodTerm{\Gm}{\lsig'}{\Sigr{\Gm}{n+1}{j}} \\
  \mathscr{P}(\lsig') &=  \top \quad \text{ when } \goodTerm{\Gm}{\lsig'}{\Sigr{\Gm}{0}{j}}  \\
\end{align*}

@@

With the above two helpers, we can define denotation $\denotesT{}$.

\begin{align*}
  \denotesT{\goodSigU{j}{\Gamma}{\lsig}{n}} &\in \setc{t}{ \goodTerm{\Gm}{t}{\Sigr{\Gm}{n}{j}} } \\
  {\goodType{\Gamma}{\TyTkg{\lsig}}{}} &= \mathscr{P}(\denotesT{\lsig}) \\
  \denotesT{\goodType{\Gamma}{\TyLkg{\sigma}}{}} &\text{ defines upon } \denotesT{\lsig} 
  \text{ and inductively on the signature length} \\
  \denotesT{\goodType{\Gamma}{\TyLkg{\sigma}}{}} &=
  \denotesT{\TyLkg{\lsigproj{1}{\sigma}}} \times \Pi(\El{\fst{(\snd{\denotesT{\sigma}})}})(\El{\app{"snd"^3~\denotesT{\sigma}}}) \\
  &\text{given } \denotesT{\goodSig{\Gamma}{\sigma}{n+1}} \\
  \denotesT{\goodType{\Gamma}{\TyLkg{\sigma}}{}} &= \top \quad
  \text{given } \denotesT{\goodSig{\Gamma}{\sigma}{0}} \\
  \denotesT{\goodSigU{j}{\Gamma}{\LSigEmp}{0}} &= \unit \\
  \denotesT{\goodSigU{j}{\Gamma}{\LSigAdd{\sigma}{f}{T}}{n+1}} &= (\denotesT{\sigma}, \codety{A}, \lam{f}, \lam{\codety{T}}) \\
  \denotesT{\goodTerm{\Gamma}{\LkgEmp}{\TyLkg{\LSigEmp}}} &= \unit \\
  \denotesT{\goodTerm{\Gamma}{\mu^+~m~t}{\TyLkg{\LSigAdd{\sigma}{s}{T}}}} &= (\denotesT{m}, \lam{t}) \\
  \denotesT{\goodTerm{\Gamma}{\Tkg{m}}{\TyTkg {\sigma}}} & \text{ is defined upon } \denotesT{m} \text{ and inductively on the signature length} \\
  \denotesT{\goodTerm{\Gamma}{\Tkg {m}}{\TyTkg {\sigma}}} &= () \quad \text{given } \denotesT{\goodSig{\Gamma}{\sigma}{0}} \\
  \denotesT{\goodTerm{\Gamma}{\Tkg {m}}{\TyTkg {\sigma}}} &= 
  (\denotesT{\Tkg {o}}, t[(\SubstExt{p^1}{f})][(\SubstExt{"id"}{\denotesT{\Tkg {o}}})]) \\
  & \text{given } \denotesT{\goodSig{\Gamma}{\sigma}{n+1}}, \text{ where } o = \lsigproj{1}{m}, t = \app{\snd{\denotesT{m}}}, f = \lsigproj{s}{\lsig}
\end{align*}
% \begin{align*}
%   \denotesT{\goodInh{\Gamma}{\_}{\sigma_1}{\sigma_2}} &= \goodTerm{\Gamma, \TyLkg{\sigma_1}}{\_}{\TyLkg{\sigma_2[p^1]}} \\ \text{ and thus } \denotesT{\goodInh{\Gamma}{h}{\sigma_1}{\sigma_2}} &\iff  \goodTerm{\Gamma, \TyLkg{\sigma_1}}{h}{\TyLkg{\sigma_2[p^1]}} \\
%   \denotesT{\goodInh{\Gamma}{"inhinh"~h~T~\uparrow^s}{(\LSigAdd {\sigma_1} {s_1}{T})}{(\LSigAdd {\sigma_2} {s_2} {T[(p^1, \uparrow^s)]})}} &= \mu^+~(h[(p^1,\lkgproj{1}{"var"_0})])\\ &\quad \quad \{f_2[p^1]\}~(\lkgproj{2}{"var"_0})[(p^1, \uparrow^s[{p^1}^{\uparrow}])] \\
% \end{align*}

>>>>

We will first define two helpers
\begin{itemize}
  \item A type $\goodType{\Gamma}{\Sigr{\Gamma}{n}{j}}{}$ where
  $\goodCtx{\Gamma}{}$ and $n \in \mathbb{N}$.
  The idea is that $\goodTerm{\Gm}{\denotesT{\goodSigU{j}{\Gamma}{\lsig}{n}}}{\Sigr{\Gm}{n}{j}}$
  \item A function $\mathscr{P} : \setc{\lsig'}{ \goodTerm{\Gm}{\lsig'}{\Sigr{\Gm}{n}{j}} } \to \setc{T}{ \goodType{\Gm}{T}{} }$ mapping a term of type $\Sigr{\Gm}{n}{j}$ to a type.
\end{itemize}
They are defined mutually inductively on $n$, the signature length.
\begin{align*}
  {\Sigr{\Gamma}{(n+1)}{j}} &= 
    \Sigma~(\Sigr{\Gamma}{n}{j})
          ~(\Sigma~\cU_j \\
          & \quad \quad ~(\Sigma~(\TyPi{{\mathscr{P}({\sub{\var{0}}{\SubstWeak{1}}})}}{\El{\sub{\var{0}}{\SubstWeak{1}}}}) \\
          & \quad \quad \quad \TyPi{\El{\sub{\var{0}}{\SubstWeak{1}}}}{\cU_j})) \\
  \Sigr{\Gamma}{0}{j} &= \top \\
  \mathscr{P}(\lsig') &=  
    \Sigma~(\fst{\lsig'})~(\El{\app{"snd"^3~\lsig'}[(\SubstWeak{1}, \app{\fst{("snd"^2~{\lsig'})}})]}) \\ & \text{ when } \goodTerm{\Gm}{\lsig'}{\Sigr{\Gm}{n+1}{j}} \\
  \mathscr{P}(\lsig') &=  \top \quad \text{ when } \goodTerm{\Gm}{\lsig'}{\Sigr{\Gm}{0}{j}}  \\
\end{align*}

@@

With the above two helpers, we can define denotation $\denotesT{}$.
%
\begin{align*}
  \denotesT{\goodSigU{j}{\Gamma}{\lsig}{n}} &\in \setc{t}{ \goodTerm{\Gm}{t}{\Sigr{\Gm}{n}{j}} } \\
  {\goodType{\Gamma}{\TyTkg{\lsig}}{}} &= \mathscr{P}(\denotesT{\lsig}) \\
  \denotesT{\goodType{\Gamma}{\TyLkg{\sigma}}{}} &\text{ is defined upon } \denotesT{\lsig}
  \text{ and inductively on the signature length} \\
  \denotesT{\goodType{\Gamma}{\TyLkg{\sigma}}{}} &=
  \denotesT{\TyLkg{\lsigproj{1}{\sigma}}} \times \Pi(\El{\fst{(\snd{\denotesT{\sigma}})}})(\El{\app{"snd"^3~\denotesT{\sigma}}}) \\
  &\text{given } \denotesT{\goodSig{\Gamma}{\sigma}{n+1}} \\
  \denotesT{\goodType{\Gamma}{\TyLkg{\sigma}}{}} &= \top \quad
  \text{given } \denotesT{\goodSig{\Gamma}{\sigma}{0}} \\
  \denotesT{\goodSigU{j}{\Gamma}{\LSigEmp}{0}} &= \unit \\
  \denotesT{\goodSigU{j}{\Gamma}{\LSigAdd{\sigma}{f}{T}}{n+1}} &= (\denotesT{\sigma}, \codety{A}, \lam{f}, \lam{\codety{T}}) \\
  \denotesT{\goodTerm{\Gamma}{\LkgEmp}{\TyLkg{\LSigEmp}}} &= \unit \\
  \denotesT{\goodTerm{\Gamma}{\mu^+~m~t}{\TyLkg{\LSigAdd{\sigma}{s}{T}}}} &= (\denotesT{m}, \lam{t}) \\
  \denotesT{\goodTerm{\Gamma}{\Tkg{m}}{\TyTkg {\sigma}}} & \text{ is defined upon } \denotesT{m} \text{ and inductively on the signature length} \\
  \denotesT{\goodTerm{\Gamma}{\Tkg {m}}{\TyTkg {\sigma}}} &= () \quad \text{given } \denotesT{\goodSig{\Gamma}{\sigma}{0}} \\
  \denotesT{\goodTerm{\Gamma}{\Tkg {m}}{\TyTkg {\sigma}}} &= 
  (\denotesT{\Tkg {o}}, t[(\SubstExt{p^1}{f})][(\SubstExt{"id"}{\denotesT{\Tkg {o}}})]) \\
  & \text{given } \denotesT{\goodSig{\Gamma}{\sigma}{n+1}}, \text{ where } o = \lsigproj{1}{m}, t = \app{\snd{\denotesT{m}}}, f = \lsigproj{s}{\lsig}
\end{align*}
% \begin{align*}
%   \denotesT{\goodInh{\Gamma}{\_}{\sigma_1}{\sigma_2}} &= \goodTerm{\Gamma, \TyLkg{\sigma_1}}{\_}{\TyLkg{\sigma_2[p^1]}} \\ \text{ and thus } \denotesT{\goodInh{\Gamma}{h}{\sigma_1}{\sigma_2}} &\iff  \goodTerm{\Gamma, \TyLkg{\sigma_1}}{h}{\TyLkg{\sigma_2[p^1]}} \\
%   \denotesT{\goodInh{\Gamma}{"inhinh"~h~T~\uparrow^s}{(\LSigAdd {\sigma_1} {s_1}{T})}{(\LSigAdd {\sigma_2} {s_2} {T[(p^1, \uparrow^s)]})}} &= \mu^+~(h[(p^1,\lkgproj{1}{"var"_0})])\\ &\quad \quad \{f_2[p^1]\}~(\lkgproj{2}{"var"_0})[(p^1, \uparrow^s[{p^1}^{\uparrow}])] \\
% \end{align*}
